From Questions to Queries – Enhancing the Curt System by a Theory and Implementation of Embedded WH-Questions

### 2008-08-29 thesis/paper.tex
A lot of work

\documentclass[11pt,a4paper]{article}

@@

\usepackage{amssymb,qtree,natbib,aleks,linguex}

\usepackage{amsthm}

@@

\usepackage{mathpazo}
% font specifications (don't overuse ;-)
\defaultfontfeatures{Mapping=tex-text} % converts LaTeX specials (``quotes'' --- dashes etc.) to unicode
\setromanfont [Ligatures={Common,Rare}, BoldFont={Fontin Bold}, ItalicFont={Fontin Italic}]{Fontin Regular}
\setsansfont [Ligatures={Common}, BoldFont={Fontin Sans Bold}, ItalicFont={Fontin Sans Italic}]{Fontin Sans}

@@

\newcommand{\abbr}{\textsf} % how to render abbreviations like IL or Ty2
\newcommand{\stress}{\textbf} % put stress somewhere (as opposed to \emph)
\newcommand{\term}[1]{\textsf{\textbf{#1}}} % introduce a term
\newcommand{\code}{\texttt} % layout of in-line code snippets
\newcommand{\pn}{\textsf} % how to render proper names like Shrdlu
\newcommand{\url}[1]{\code{http://#1}} % how urls appear
\newcommand{\example}{\textit} % in-text examples

@@

@@

@@

Growing efforts have thus been made to provide a technology that is (going to
be) able to extract logical formulae from sentences of ordinary language and
establish relations between objects of its logical representation. In addition,
providing a means to compute with said representations will eventually enable
the machine-processing of an abstract concept of meaning. In doing so, the
ultimate goal may be to enable a man-made system to \emph{understand} what a
humans communicate.

Exploration of this  interesting field of linguistics has shown that
the notion of \emph{understanding} is far more than a linguistic concern, but
involves at least findings in the domain of logics, (cognitive) psychology and
probably even philosophy.

@@

As the scope of a Bachelor's thesis is limited by its nature, not all of the
theory and reasoning can be understood without background knowledge in some
rather divergent domains. In particular, a basic familiarity with traditional
\term{natural language semantics} and \term{extensional} as well as
\term{intensional logic} frameworks will be assumed. Furthermore, a basic
acquaintance with \term{eroteric logic} will be helpful in understanding some of
the reasoning in section \ref{sec:theory}.
Some knowledge of the \prol{} programming language and a good overview of the
architecture of the \curt{} system is imperative in order to understand the
implementation itself.

@@

@@

Computational Semantics is a discipline that has  lately gained a wider
focus in literature due to recent progress in different areas.
Current development in computer hardware has made it possible to focus on
theoretical soundness and important abstract designs and concepts. On the other
hand, research in both computer science and linguistics has fueled new ideas and
intriguing hypotheses emerge from the discussion on a very frequent basis.

This section delineates the ground the software presented in this thesis is
based on. \ref{sec:curt} introduces the base system of the implementation, \curt
and \ref{sec:comparison} relates it to other similar systems in existence.
\ref{sec:extension} then goes on to explain where the \curt{} system lacks in
features and thus exhibits potential for extension.

@@

\curt, a system written in \prol{} that is able to interpret basic utterances
of English and store them in expressions of first order logic, is
presented in \cite[chapter 6]{blackburnbos:cl1}. While relatively limited when
compared to other systems presented in \ref{sec:comparison} in terms of lexical
and grammatical coverage, it provides a sound basis for extended work because of
its coherent design and relative simplicity.

@@

The final chapter of \cite{blackburnbos:cl1} introduces the implementation
incrementally, presenting new versions of \curt{} that add to the software's
feature set. Support for external inferencing via theorem provers and model
builders is added right after the first naïve implementations and
finally an ontology and a straight-forward way of parsing and answering direct
questions for noun phrases conclude its development. The theory and
implementation of this paper is meant as a replacement and enhancement for the
latter feature. This will also happen in an incremental manner and new features
will be introduced one at a time.

@@

Generally, question answering and semantic parsing are not related and seldom
interdependent in current implementations of similar systems, with a few notable
exceptions. As question answering becomes a more and more sophisticated task,
the systems employed gather an increasing capability of \emph{understanding}
their (input) domain. On the other side of the fence, interest in reasoning
tools and inference engines operating on natural language have grown
considerably over the past decade and some have stretched out to encompass more
advanced domains of logic and discourse representation.

\cite{chat} are among  the first to attempt creating a question answering system
with deep semantic coverage. Their effort has been focused on the geographical
domain, and their implementation, dubbed \pn{CHAT-80}, is written entirely in
\prol. The lack of hardware resources of that time forced the system to be
written with efficiency in mind. Moreover, it had to maintain a statical
database of facts which was hand-crafted. The \curt{} system is built on  an
entirely different set of premises and aims to provide a proof-of-concept of a
more general purpose system that is also able to extend its knowledge
dynamically.

Soon after \pn{CHAT-80}, two entirely different types of systems emerged. The
first type is probably the most promising to be used on a wider scale, given
today's technological standards. Implementations are usually based on
\term{information retrieval} and the generation of question/answer pairs. Such
systems are generally referred to as a \abbr{QA} (\term{Question Answering})
systems.  Examples implementations of such systems are discussed in \cite{rama},
\cite{iesqa} and many others. Most of these seem to provide a good lexical and
grammatical coverage due to their large input corpora and
shallow interpretation techniques. The latter is also the \abbr{QA} technique's
greatest shortcoming, as no deep analysis is  performed and usually
only factoids of less then a couple of words in length can be supplied as
answers.\footnote{But see \cite{aqabtf} for an approach that aims to cover a
wider range of possible answers.}

In contrast, \pn{CHAT-80} runs all its queries against a pre-built database
without the ability to acquire new knowledge from the discourse. While a
\abbr{QA} implementation will typically just acquire knowledge by consuming
more input and analysing its statistical properties, programs employing deep
semantic analysis will have to check their input for logical contradictions in
to maintain their database in a reasonable\footnote{FIXME pun intended, but
appropriate? %todo
} state.

\cite{prontoqa} introduce the \pn{Pronto QA System}, which couples traditional
corpus-based \abbr{QA} techniques with deep semantic parsing based on
\term{Discourse Representation Theory}. It is one of the few \abbr{QA} systems
with a capability of reasoning about its input text. Unfortunately, there seems
to be no working demo to date. The homepage of the ongoing efforts of Johan Bos
and his team can be found at the following address:
\href{http://svn.ask.it.usyd.edu.au/trac/candc/}{\url{http://svn.ask.it.usyd.edu.au/trac/candc/}}.

\pn{Godot}, a very interesting project, is introduced in \cite{godot}.
\pn{Godot} is a so-called \term{embodied agent}, a robot with a sophisticated
software driving interaction with the outer world. As such its conception of
discourse not only relies on spoken dialogue, but also on sensory input.
\pn{Godot} is focused on executing orders -- after evaluating their feasibility
in its particular discourse model -- and does so far not seem to support the answering
of questions about its state.

@@

\begin{termin}
  An \term{interrogative} is a certain type of sentence in natural language
  distinguished by certain characteristics\footnote{To quote \cite{gs:q}: \quote{[An
  interrogative sentence is] characterized by word order, intonation, question
  mark, the occurrence of interrogative pronouns}.}
  whereas an \term{interrogative act} refers to the utterance of an
  \term{interrogative}. Finally, a
  \term{question} will denote the semantic entity uttered by the
  \term{interrogative act}, as it will be introduced in section
  \ref{sec:protoq}
\end{termin}

@@

\begin{quote}A direct question can be treated as semantically equivalent to a
certain kind of declarative sentence containing the corresponding indirect
question embedded under  a suitable `performative' verb.\end{quote}

Using this strategy, Karttunen escapes the treatment of direct  questions and
thus gives an approach for indirect questions only.
In this respect, his analysis is in certain aspects similar to the one presented in
\cite{gs:sawhq}, but the latter goes on to challenge certain aspects of his
hypothesis.

@@

\begin{quote}
  Questions, unlike assertions, don't have truth-values, so it would be
  misleading to represent them as ordinary formulas.
\end{quote}

Their notion of questions will be extended for \acurt in order to account for
yes/no interrogatives and larger sets of answers made possible by using model
which allow for possible worlds.

Another important distinction in the definition of questions, is the definition
of \term{answerhood}, or, what constitutes a valid answer to a given
interrogative. The ground-laying work by \cite{hamblin:q} states that in knowing
all possible answers to a question, one also knows the question itself. The
notion of answerhood is going to be elaborated on in section \ref{sec:answerhood}

One thing remains clear though, no matter what the notion of a question or
finding an answer to it is; upon
finding an answer, said answer has to be \emph{realized} in natural language
using text or speech synthesis. When realizing an answer, many pragmatic
considerations have to be taken into account.
However, as the uttering of an embedded question does not prompt an
answer\footnote{Unless, of course, the question is uttered as an
\term{imperative} -- but this would pose an entirely different class of problem
that is not discussed here.}, the present
work will not directly concern itself with a solution to this problem.

@@

\begin{enumerate}
  \item NP-questions ask for a noun phrase, e.g. \example{Who shot Vincent?
  Butch} or even more then one noun phrase at the same time \example{Who had
  what in the hash bar?}
  \item PP-questions ask for prepositional constituent, e.g. \example{Where did
  Jules stay? In Amsterdam.}
  \item VP-questions ask for an action performed by an individual, e.g.
  \example{What did Butch do? He shot Vincent.}
  \item Temporal questions ask for the date of a certain event, e.g.
  \example{When did Butch shoot Vincent? When he entered the toilet.}
  \item etc\ldots
\end{enumerate}

This classification of interrogatives can go on for quite some more entries,
including temporal, modal, causal and all sorts of other questions. Out of all
those categories, only NP-questions will be supported, with an option to
integrate other sorts of questions at a later stage.

@@

\pn{Helpful Curt}, the most advanced version of \curt{} \cite{blackburnbos:cl1}
present, only addresses the problem of direct constituent interrogatives and
does neither treat yes/no interrogatives nor embedded questions. \acurt treats
embedded questions by embedding a special construct, a \term{proto-question},
under verbs of certain types (which are going to be discussed in
\ref{sec:ontology}). Furthermore, the yes/no interrogative is considered a
special case of the constituent interrogative.

\begin{definition}\label{def:qop}
  \stress{The Question Operator}
  \begin{enumerate}
    \item If $\varphi$ is a proposition in which the free variables
    $x_1,\ldots,x_n$ for arbitrary values of $n \geq 1$ occur,
    $?\pair{x_1,\ldots,x_n}.\varphi$ denotes a \term{question}, associated with
    a constituent interrogative.
    \item If $\varphi$ does not contain any free variables, $?.\varphi$ is
    considered a question associated wit a yes/no interrogative.
    \item The question operator does \emph{not combine} with any other operator.
  \end{enumerate}
\end{definition}

@@

Note that the last clause of definition \ref{def:qop} enforces the question operator to
always be the last operator of a formula of propositional logic. Embedded
questions will thus not have to operate on the questions directly, but rather on
their value (answers), or the set of propositions they denote.

As was already mentioned, \pn{Helpful Curt} already has a certain albeit limited
support for evaluating NP constituent interrogatives. Blackburn and Bos adopt a
sort of pseudo-notation of the question operator in what they call an
extra-logical form which does not itself a yield a truth value, but is
internally evaluated in a special case.

They treat a question as a sort of \emph{quantification
over a certain individual $X$, for which $R$ and $S$ must hold.} Thus a parse of
a sentence like \ref{ex:bbqueexa} and \ref{ex:bbqueexb} would yield the representation given in
\ref{ex:bbqueparsea} and \ref{ex:bbqueparseb}.

@@

The aforementioned special case in \pn{Helpful Curt} only evaluates a question
against the extensional model which is used as a representation of \curt's state
and knowledge. In this it differs substantially from the approaches made by
Groenendijk and Stokhof and also later approaches by \cite{g:is}. \acurt will
bridge the gap between the two.

@@

from questions...
                   Hamblin’s postulates read as follows:
     (i) An answer to a question is a sentence, or statement.
    (ii) The possible answers to a question form an exhaustive set of mutually
exclusive possibilities.
   (iii) To know the meaning of a question is to know what counts as an answer to
that question.

@@

An important observation to be made here is that only \emph{assertive} verbs can
embed real questions. \emph{Stipulating} verbs only embed \emph{propositions} as
in \ref{ex:stipul}.
Nevertheless, both are treated in the following analysis, but with a focus on
the treatment of assertive \wh-embedding verbs.

\subsubsection{Modeling Knowledge}\label{sec:epistemic}

>>>>

\documentclass[a4paper]{article}

@@

\usepackage{amssymb,qtree,natbib,bbm,aleks,linguex}

\usepackage{stmaryrd,amsthm,amsmath}

@@

\usepackage{mathpazo}
% font specifications (don't overuse ;-)
\defaultfontfeatures{Mapping=tex-text} % converts LaTeX specials (``quotes'' --- dashes etc.) to unicode
\setromanfont [Ligatures={Common,Rare}, BoldFont={Fontin Bold}, ItalicFont={Fontin Italic}]{Fontin Regular}
\setsansfont [Ligatures={Common}, BoldFont={Fontin Sans Bold}, ItalicFont={Fontin Sans Italic}]{Fontin Sans}
\setmonofont [Ligatures={NoCommon,NoRequired,NoRare}] {Monaco}

@@

\newcommand{\abbr}{\textsf} % how to render abbreviations like IL or Ty2
\newcommand{\stress}{\textbf} % put stress somewhere (as opposed to \emph)
\newcommand{\term}[1]{\textsf{\textbf{#1}}} % introduce a term
\newcommand{\code}[1]{\footnotesize\texttt{#1}\normalsize} % layout of in-line code snippets
\newcommand{\pn}{\textsf} % how to render proper names like Shrdlu
\newcommand{\url}[1]{\code{http://#1}} % how urls appear
\newcommand{\example}{\textit} % in-text examples

@@

%quotations
\newenvironment{quotes}{\begin{quote}\sf}{\rm\end{quote}}
\newcommand{\quotesf}[1]{\quote{\textsf{#1}}}

@@

\theoremstyle{definition}
\newtheorem{postulate}[thm]{Postulate}

@@

Growing efforts have thus been made to provide a technology that is (going to
be) able to extract logical formulae from sentences of ordinary language and
establish relations between objects of their logical representation. In addition,
providing a means to compute with said representations will eventually enable
the machine-processing of an abstract concept of meaning. In doing so, the
ultimate goal may be to enable a man-made system to \emph{understand} what
humans communicate.

Exploration of this  interesting field of linguistics has shown that
the notion of \emph{understanding} is far more than a linguistic concern, and
involves at least findings in the domain of logics, (cognitive) psychology and
probably even philosophy.

@@

As the scope of a Bachelor's thesis is limited by its nature, not all of the
theory and reasoning can be understood without background knowledge in some
rather divergent domains. In particular, a basic familiarity with traditional
\term{natural language semantics} and \term{extensional} as well as
\term{intensional logic} or \term{modal} frameworks will be assumed.
Furthermore, a basic acquaintance with \term{eroteric logic} will be helpful,
but not strictly necessary in understanding some of the reasoning in section
\ref{sec:theory}.  Some knowledge of the \prol{} programming language and a good
overview of the architecture of the \curt{} system is imperative in order to
understand the implementation itself.

@@

\subsection{Typographic Conventions}

The following typographic conventions are used throughout the document to
increase readability:

\begin{itemize}
  \item Text in a bold face sans serif font marks new terminology and important
  concepts: \term{embedded interrogatives}, \term{backtracking}
  \item Proper names, names of technologies, and abbreviations are typeset in a
  regular sans serif font: \acurt, \prol, \pn{Ty2}
  \item Code is written in a monospace font, resembling the look of a font class most
  programmers are used to:\\
  \code{lam(P, lam(Q, all(X, imp(app(P,X),app(Q,X)))))}.
  \item Logical formulae are typeset in a different serif font from the rest of
  the text:\\
    Assume $k = (\lambda x.\neg xx)$. Then $ kk = (\lambda x .\neg xx)k = \neg kk$
  \item Quotations  are typeset on a separate paragraph, using a sans serif font:
  \quotesf{This answers the question!}
\end{itemize}

@@

Computational Semantics is a discipline that has  lately gained a wider focus in
literature due to recent progress in different areas.  Current development in
computer hardware has made it possible to focus on theoretical soundness and
important abstract designs and concepts of logic and the theory of meaning
rather than being able to simply implement a system that is able to do basic
computations in a reasonable amount of time. On the other hand, research in both
computer science and linguistics has fueled new ideas and intriguing hypotheses
emerge from lively discussion in their respective fields on a very frequent
basis.

This section delineates the ground the software presented in this thesis is
based on. \ref{sec:curt} introduces the base system of the implementation,
\curt{}
and \ref{sec:comparison} relates it to other similar systems in existence.
\ref{sec:extension} then goes on to explain where the \curt{} system lacks in
features and thus exhibits potential for extension.

@@

\curt, a system written in \prol{} that is able to interpret basic utterances
of English and store them in expressions of first order logic, is
presented in \cite[chapter 6]{blackburnbos:cl1}. While relatively limited when
compared to other systems presented in \ref{sec:comparison} in terms of lexical
and grammatical coverage, it provides a sound basis for extended work due to
its coherent design and relative simplicity.

@@

The final chapter of \cite{blackburnbos:cl1} introduces the implementation
incrementally, presenting new versions of \curt{} that add to the software's
feature set one at a time. Support for inferencing via external theorem provers
and model builders is added right after the first, rather limited implementations.
Finally an ontology and a straight-forward way of parsing and answering direct
questions for noun phrases conclude its development. The theory and
implementation of this paper is meant as a replacement and enhancement for the
latter feature. This will also happen in an incremental manner and new features
are introduced one at a time during section \ref{sec:implementation}.

@@

Generally, software that concerns itself with answering questions formulated in
ordinary human language falls in two main categories: \term{information
retrieval systems}, also known as \abbr{QA} (question answering) systems, and
\term{dialogue systems}.

The purpose of the former is in most cases twofold, although present
\abbr{QA} systems vary in the functionality they offer
and results they actually deliver. One goal is to enable their users to
query for information without the need for a special query language or at least
using a query language that would be easy and natural to use even for the
uninitiated user. This way the user can pose direct questions to a system
instead of having to search for a keyword to occur in the plain text of a
document, like they would have to do with traditional information
retrieval systems. The second feature provides \abbr{QA} systems with another important
distinction, viz. the ability to utilize ontologies in order to relate query
terms akin, but not equal to words in target documents and thus deliver
satisfactory results even for fuzzy queries.

These systems typically do not try to \emph{reason} about
their input, but gain `knowledge' from applying statistical methods on large
input corpora. The methods used and results achieved vary throughout the
literature. One technique employed is feeding a database with large sets of
question/answer pairs, as described in \cite{rama} and then running basic
natural language processing tasks like part-of-speech tagging,
named-entity-recognition, syntactic and semantic parsing on the input.

Many systems are only able to deliver small bits of information, also known as
\term{factoids}, or just display fragments of their input text as it occurred in
their corpus. Their knowledge base is typically not interactively modifiable,
and acquiring knowledge and answering questions are two entirely separated
tasks.

\cite{prontoqa} introduce the \pn{Pronto QA System}, which couples traditional
corpus-based \abbr{QA} techniques with deep semantic parsing based on
\term{Discourse Representation Theory}. Unfortunately, there seems
to be no working demo to date. The homepage of the ongoing efforts of Johan Bos
and his team can be found at the following address:
\href{http://svn.ask.it.usyd.edu.au/trac/candc/}{\url{http://svn.ask.it.usyd.edu.au/trac/candc/}}.

Interestingly, earlier attempts at providing a software capable of answering
questions in natural language where also mostly coupled with the usage of
semantic analysis instead of statistical methods.
\cite{chat} are among  the first to attempt creating a question answering system
with reasonable semantic coverage. Their effort has been focused on the geographical
domain, and their implementation, dubbed \pn{CHAT-80}, is written entirely in
\prol. The lack of hardware resources of that time forced the system to be
written with efficiency in mind. Moreover, it had to maintain a statical
database of facts which was hand-crafted. As such, it falls more into the
category of question answering systems than dialogue systems.

Today it seems that question answering systems make inroads to the search market,
and some argue that they are bound to eventually supplant current document/term-based
search engines. Even though \abbr{QA} systems also operate on the domain of
answering questions, their approach to resolving them differs in most aspects
from that of dialogue systems. The latter have a much higher obligation to
correctness, since their results are typically not subject to peer
review and server an entirely different purpose.

\pn{Godot}, a very interesting project, is introduced in \cite{godot}.
\pn{Godot} is a so-called \term{embodied agent}, a robot with a sophisticated
software driving interaction with the outer world. As such its conception of
discourse not only relies on spoken dialogue but also also on sensory input, rather
than on an input corpus. Its knowledge is also \emph{dynamic}, i.e. it can be
extended through the aforementioned means, making it
dialogue system.
\pn{Godot} is focused on executing orders -- after evaluating their feasibility
in its particular discourse model -- and does so far not seem to support the answering
of questions about its state.

\curt{} is a dialogue system and thus based on an entirely different set of
premises than \abbr{QA} systems. Moreover, it differs from \pn{Godot} in some
major points, the most distinctive of which is that \curt{} only exists inside a
regular computer and has no access to sensory input. Also, \curt's
representation of knowledge differs from that of \pn{Godot} or the \pn{Pronto QA
System} in that it does not use discourse representation theory (yet). \acurt{}
addresses some of its shortcomings and lays ground to the implementation of a
broader feature set for it.

@@

\begin{termin}
  An \term{interrogative} is a certain type of sentence in natural language
  distinguished by certain characteristics
  \footnote{To quote \cite{gs:q}: \quotesf{[An
  interrogative sentence is] characterized by word order, intonation, question
  mark, the occurrence of interrogative pronouns}.}
  whereas an \term{interrogative act} refers to the utterance of an
  \term{interrogative}. Finally, a
  \term{question} will denote the semantic entity uttered by the
  \term{interrogative act}, as it will be introduced in section
  \ref{sec:protoq}
\end{termin}

@@

\begin{quotes}A direct question can be treated as semantically equivalent to a
certain kind of declarative sentence containing the corresponding indirect
question embedded under  a suitable `performative' verb.\end{quotes}

Using this strategy, Karttunen escapes the treatment of direct  questions and
thus gives an approach for indirect (i.e. \emph{embedded}) questions  only.
In this respect, his analysis is somewhat similar to the one presented in
\cite{gs:sawhq}, but the latter goes on to challenge certain aspects of his
hypothesis.

@@

\begin{quotes}
  Questions, unlike assertions, don't have truth-values, so it would be
  misleading to represent them as ordinary formulas.
\end{quotes}

Their notion of questions will be extended for \acurt{} in order to account for
yes/no interrogatives and larger sets of answers made possible by using model
which allow for possible worlds.

@@

\begin{enumerate}
  \item NP-questions ask for a noun phrase, e.g. \example{Who shot Vincent?
  Butch (shot Vincent.)} or even more then one noun phrase at the same time, \example{Who had
  what in the hash bar?\footnote{Multiple constituent interrogatives pose a
  somewhat greater challenge, since they come with an inherent ambiguity, the
  so-called \term{Baker Ambiguity}, as presented in \cite{baker:1970}.
  Such ambiguities will not be concerned here.}}
  \item PP-questions ask for prepositional constituent, e.g. \example{Where did
  Jules stay? In Amsterdam.}
  \item VP-questions ask for an action performed by an individual, e.g.
  \example{What did Butch do? He shot Vincent.}
  \item Temporal questions ask for the date of a certain event, e.g.
  \example{When did Butch shoot Vincent? When he entered the toilet.}
  \item etc\ldots
\end{enumerate}

This classification of interrogatives can go on for quite some more entries,
including local, modal, causal and all sorts of other questions. Out of all
those categories, only NP-questions will be supported, with an option to
integrate other types of interrogatives at a later stage.

Another important distinction in the definition of questions, is the definition
of \term{answerhood}, i.e. what constitutes a valid answer to a given
interrogative. The ground-laying work by \cite{hamblin:q} states that in knowing
all possible answers to a question, one also knows the question itself. The
notion of answerhood is going to be elaborated on in section \ref{sec:answerhood}

One thing remains clear though, no matter what the notion of a question or
finding an answer to it is; upon
finding an answer, said answer has to be \emph{realized} in natural language
using text or speech synthesis. When formulating an answer, many pragmatic
considerations have to be taken into account.
However, as the uttering of an embedded question does not prompt an
answer\footnote{Unless, of course, the question is uttered as an
\term{imperative} or \term{performative}-- but this is an entirely different
class of problems that is not discussed here.}, the present
work will not directly concern itself with a solution to this problem.

@@

\pn{Helpful Curt}, the most advanced version of \curt{} \cite{blackburnbos:cl1}
present, only addresses the problem of direct constituent interrogatives and
does neither treat yes/no interrogatives nor embedded questions. The
interrogative operator, as it will be introduced shortly, is designed to cover
the semantics of both types of questions simultaneously. It takes a proposition
(a formula yielding a truth value as a result) but gives back a set of
propositions. Computation on more than one proposition in parallel will be
introduced into the \curt{} systems in section \ref{sec:possworlds}.  \acurt{}
represents embedded questions by embedding a special construct, a
\term{proto-question}, under verbs of certain types (which will be discussed
later on). Since the yes/no interrogative is considered a special case of the
constituent interrogative the semantic formulae produced by the grammatical
back end of \acurt{} do not differ for the two cases.

\begin{definition}\label{def:qop}
  \stress{The Question Operator}
  \begin{enumerate}
    \item If $\varphi$ is a proposition in which the free variables
    $x_1,\ldots,x_n$ for arbitrary values of $n \geq 1$ occur,
    $?\pair{x_1,\ldots,x_n}.\varphi$ denotes a \term{question}, associated with
    a constituent interrogative.
    \item If $\varphi$ does not contain any free variables, $?.\varphi$ is
    considered a question associated wit a yes/no interrogative.
    \item \label{def:outermost}The question operator does \emph{not combine} with any other operator.
  \end{enumerate}
\end{definition}

@@

Note that the last clause of definition \ref{def:qop} enforces the question operator to
always be the last operator of a formula. This constraint will be overridden
during parsing of questions to produce forms that do have nested questions.
When calculating the models and readings of a formula, however, these operators
will be resolved first. This is done in order to increase the efficiency during
parsing, since the syntactic parser will sometimes have to backtrack.

As was already mentioned, \pn{Helpful Curt} has a certain albeit limited
support for evaluating NP constituent interrogatives. Blackburn and Bos adopt a
sort of pseudo-notation of the question operator in what they call an
extra-logical form which does not itself yield a truth value, but is
internally evaluated in a special case.

They treat a question as a \emph{quantification
over a certain individual $X$, for which $R$ and $S$ must hold.} Thus a parse of
a sentence like \ref{ex:bbqueexa} and \ref{ex:bbqueexb} would yield the representation given in
\ref{ex:bbqueparsea} and \ref{ex:bbqueparseb}.

@@

The aforementioned special case in \pn{Helpful Curt} only evaluates a question
against the extensional model which is used as a representation of \curt's state
and knowledge. In this it differs substantially from the approaches made by
Groenendijk and Stokhof and also later approaches by \cite{g:is}. \acurt{}
bridges the gap between both worlds.

@@

Knowing the structure of an interrogative, the next step is to evaluate the
question's \stress{value}, namely the propositions it generates. This will
invariably bring up the question of the definition of \term{answerhood}, viz.
the relation of the question to its answer or answers with respect to the
current discourse. It is important to note that an answer other than `no'
`nobody' cannot occur without a certain discourse or background knowledge already being present
for the answer to be evaluated against.

\cite{hamblin:q} defines a basic notion of answerhood, relating the value of
a question with the set of propositions that might serve as partial answers to
it. \cite{gs:q} phrase Hamblin’s postulates as follows:

\begin{quotes}
  \begin{enumerate}
    \item An answer to a question is a sentence, or statement.
    \item The possible answers to a question form an exhaustive set of mutually
    exclusive possibilities.
    \item To know the meaning of a question is to know what counts as an answer to
    that question.
  \end{enumerate}
\end{quotes}

For yes/no interrogatives this means, that if the question is $?.\varphi$, the
answer two it is either $\varphi$ or $\neg\varphi$. For constituent
interrogatives, this question is not so easily answered.  There is one central
issue to be addressed first: the issue of the definition of answerhood: should
the set of answers be an \emph{exhaustive} set of answers, including the
negative ones, or just the set of possible \stress{true} answers?

The following situation might serve as an example: Suppose Mia and Vincent both
play air guitar. Butch and Marcellus do not play air guitar. When answering the
question \example{Who plays air guitar?}, one can observe three degrees of
exhaustiveness, according to \cite[section 1.5]{gs:sawhq}. Clearly, it does not
suffice to just say \example{Vincent.} or \example{Mia.}, since that would still
not account for all possible answers to the question, and that both
\example{Butch.} and \example{Marcellus.} would be outright wrong. In the
analysis of \cite{karttunen:1977}, every set of propositions that is a superset
of the set of propositions containing \example{Mia plays air guitar.} and
\example{Vincent plays air guitar.} would be a valid answer. This is the lowest
degree of exhaustiveness, and accounts for all \emph{true} answers. So, if
Yolanda believed to not only know of Vincent and Mia that they played air
guitar, but also of Butch, she would still know all possible true answers, and,
according to this analysis, a correct answer.

Groenendijk and Stokhof argue for two more degrees of exhaustiveness. The first
one would require the answer not only to contain all possible \emph{true}
propositions, but also to \emph{not contain any \stress{false} propositions}. So
in this case, if Yolanda believed Butch played air guitar, she would not know
the true answer.

The second and strongest degree of exhaustiveness would
require any set of answers not only to contain the set of propositions that
hold, but also the set of propositions that \emph{do not hold}. In this case, for
the sentence \example{Yolanda knows who plays air guitar.} to be true, the
sentence \example{Yolanda knows who does not play air guitar.} would
automatically follow.

There even more ways of categorizing the number of answers to a question. So far
the so called \term{mention-all} case has been discussed, but when considering
a sentence like \example{John knows where to buy a Chinese newspaper.} (example
due to \cite{rooy:qar}), it is clear that only one or two out of all possible
propositions for \example{Where to buy a Chinese newspaper?} will be sufficient
for John to know in order for him to know where to actually buy one. This has
been regarded as the \term{mention-some} reading.

Even after having settled on the \emph{number} and \emph{value} of propositions
that have to serve as an answer to a question, one still might be concerned with
their kind. Taken the previously given example, if Butch and Marcellus were
all boxers in this situation, would not an answer such as \example{Every boxer.}
suffice? \cite{rooy:qar} elaborates on the many aspects of different analyses
regarding the relevance of answers. Suffice to say, \acurt{} is only content
with \emph{maximally specific} resolutions to answers.

Moreover, as \acurt{} will only have to deal with answers to questions
internally, it will also only regard maximally exhaustive interpretations of
answers.
Correctly answering a direct question with respect to relevance and other
pragmatic factors is a task left as a task for turning \curt{} into a complete
dialogue system more potent of dealing with queries to its knowledge.

After having agreed on a treatment of interrogatives, the next step is to
construct a framework for embedding the result under certain verbs. The main
issues of this section will thus be focused on providing a classification of
embedding verbs and elaborating on the details of the relationship between
embedded and embedding constituents of an utterance.

As was already mentioned, clause \ref{def:outermost} of definition \ref{def:qop}
will be overridden when treating embedded interrogatives. The following example
illustrates a syntactic parse of an embedded interrogative:

\ex. \label{ex:whembed}
\a. Mia knows who snorts.\footnotesize
\b.  \Tree
[.$t$\\$know(?X_1.snort(X_1))(Mia)\wedge ?X_1.snort(X_1)$ [.$T$ { $\lambda X.X(Mia)$\\Mia } ]
[.$IV$\\$\lambda X.know(?X_1.snort(X_1))(X)\wedge ?X_1.snort(X_1)$ [.{$IV/\bar{t}$} 
{ $\lambda S.\lambda X.[  \lambda Y.[ know(Y)(X) \wedge Y ](S) ]$\\knows } ]
[.{$\bar{t}$\\$?X_1.snort(X_1)$ } [.$\bar{t}/t$ ? ] [.{ $t$\\$snort(X_1)$ }
[.$T$ { $X_1$\\who } ]
[.$IV$ { $\lambda X.snort(X)$\\snorts } ] ] ] ]!\qsetw{4cm} ]\normalsize
\b. $know(?X_1.snort(X_1))(Mia)\wedge ?X_1.snort(X_1)$\label{ex:formula}

Note that this example is using  the (syntactic) type system of \cite{gs:sawhq}
but different meaning postulates. Grammar rules, later explained in
\ref{sec:curtgrammar}, are
responsible for determining the correct order of functional application in the
untyped logical system \acurt{} is using.

One can see \ref{ex:formula} as being an \emph{intensional} denotation of the
meaning of \ref{ex:whembed} rather than an extensional one. In fact, this
sentence will evaluate to more than one proposition in case there is more than one
person snorting.

The reason for $?$ to be embeddable within more complex formulae is merely one
of convenience, since one could transform $?X.\varphi(X)$ to mean

\[\forall w \in W_M [ \forall x \in D_w [ \varphi(w)(x) = 1 ] ]\]

assuming $W_M$ is the set of possible worlds in a model $M$, but this would
require modifications to the current way of evaluating predicates with \curt{}
as well as the definition of a quantification over possible worlds, which is
currently not supported and would require a special treatment. Furthermore, in
the case that the explicit formula does not yield any results, the question
should return its correct answer

\[ \neg\exists x [\varphi(x)] \]

Instead, before calling its inferencing tools \acurt{} will look out only for
the question operator itself and take appropriate measures.

Another advantage of this approach is that the treatment for yes/no
interrogatives and constituent interrogatives will not require special
intervention at the level of grammar.

\ex. \label{ex:simpleembed}
\a. Mia knows whether Vincent snorts.  \footnotesize
\b. \Tree
[.$t$\\$know(?.snort(Vincent))(Mia)\wedge ?.snort(Vincent)$ [.$T$ { $\lambda X.X(Mia)$\\Mia } ]
[.$IV$\\$\lambda X.know(?.snort(Vincent))(X)\wedge ?.snort(Vincent)$ [.{$IV/\bar{t}$} 
{ $\lambda S.\lambda X.[  \lambda Y.[ know(Y)(X) \wedge Y ](S) ]$\\knows } ]
[.{$\bar{t}$\\$?.snort(Vincent)$} [.$\bar{t}/t$ whether ] [.{ $t$\\$snort(Vincent)$ } 
[.$T$ { $\lambda X.X(Vincent)$\\Vincent } ]
[.$IV$ { $\lambda X.snort(X)$\\snorts } ] ] ] ]!\qsetw{5cm} ]
\normalsize
\b. $know(?.snort(Vincent))(Mia)\wedge ?.snort(Vincent)$

The question operator in front of saturated formulae ($?.\varphi$) could also be
rephrased as

\[ \forall w\in W_M [ \varphi(w) \wedge \neg\varphi(w) \]

This would again demand explicit quantification over worlds and a different
treatment of predicates. Instead, it can be treated with the same procedure as
before when using an embedded question operator.

Embedded clauses are always embedded under a certain verb, which usually
takes a certain type of phrase and returns an intransitive verb. Examples of
this were given in the previous section was in \ref{ex:simpleembed} and
\ref{ex:whembed}. The verb used in both example was \example{know}. The meaning
postulate for \example{know} is the following:

\begin{postulate}\label{def:know}
  Knowledge necessitates the belief of a proposition $p$ by the knowing
  individual as well as the verity of $p$.

  \sem{know} = $\lambda p\lambda y.[know(y)(p)\wedge p]$
\end{postulate}

While the argumentation about exhaustiveness in section \ref{sec:answerhood}
already drew upon the this statement, postulate \ref{def:know} puts this into a
more formal perspective. The postulate is based on the fact that, clearly, one
can not \emph{know} something that is not true. In that case one's knowledge
would be wrong, and one would only \emph{believe to know}.

The proposition $snorts(Vincent)$, which would evaluate to a truth value, is
turned into a \wh-construct by \example{whether} -- the exact circumstances of
this will be explained further on.

@@

An important observation to be made here is that only \emph{assertive} verbs can
embed real questions. \emph{Stipulating} verbs only embed \emph{propositions} as
in \ref{ex:stipul}.
Nevertheless, both are treated in the following analysis, but with a focus on
the treatment of assertive \wh-embedding verbs.
\subsubsection{Modeling Knowledge}\label{sec:epistemic}
